Ornaments as indicators of social changes resulting from indirect effects of colonialism in northeastern Taiwan

### 2020-09-24 analysis/paper/paper.Rmd
udpate abstract and intro based on the proof version

---
title: "Ornaments as indicators of social changes resulting from indirect effects of colonialism in northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      #reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on Indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1350-1850 AD) in northeastern Taiwan show the indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered. This hints at increasing social inequality resulting from colonial influence. Ornaments give insights into the increasing social inequality stimulated by the European colonial presence, and show the agency of Indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; European colonial period; 17th century; Trade ornaments; Exchange networks; spatial pattern; East Asia
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local Indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

The direct impacts of European colonialism on Indigenous communities in East Asia were much less conspicuous than in island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local Indigenous communities remains largely unanswered. Understanding the indirect effects of colonialism are important for detecting colonial impacts on Indigenous peoples in the periphery of colonial control [@Acabado2017; @Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local Indigenous societies caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Consumption patterns of foreign goods can give insights into negotiations between colonized and colonizer, and the resistance and accommodations of Indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000; @Voss2005]. Northeastern Taiwan is an ideal context to study peripheral colonial influence because although there was a prominent Spanish and Dutch colonial presence in parts of Taiwan, the northeastern region was isolated from intensive direct contact by the Xueshan Mountains.

This article describes personal ornaments excavated from the upper component of Kiwulan (1350 AD-1850 AD), the largest Iron Age settlement on the Yilan plain in northeastern Taiwan. Ornaments are found at many Iron Age sites in Yilan, but only Kiwulan shows clear stratigraphic contexts from pre-European period to modern time. The first recorded European presence in Yilan was a Spanish revenge attack on Indigenous villages in 1632 [@Borao2001, pp. 163]. In 1647 the Dutch attacked villages and forced them to accept colonial rule and pay an annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest Indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. Following defeat of Dutch by the Chinese general Koxinga in 1661-1662, the Dutch abandoned northern Taiwan. Direct contact with Han Chinese is indicated by Qing dynasty census reports mentioning Yilan villages in 1821 [@Yao1996].

@@

Figure \@ref(fig:plot-altogether) shows the distribution of frequencies for subtypes in each major class. Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of the effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bells and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells have less variety, typically >2 cm long with a wide variety of human faces as a motif. Although glass beads have less variety in size, presenting as small (0.5-1 cm) or medium (1-2 cm), they have a wide variety of colors or patterns mostly made by a winding technique and with high lead content indicating possibly from China [@Cheng2008].

@@

An indirect colonial influence may be indicated at Kiwulan by the greater diversity of ornament types and materials during the European period. Yilan was involved in complex trading networks both on a regional scale with other Indigenous groups and Chinese merchants, and at a global scale with Europeans, including the Dutch and the Spanish. Those trade ornaments have multiple origins, including Southeast Asia and China, and were first introduced into northeastern Taiwan by Chinese merchants before the 17th century. Later, trade activities became more frequent and intense in the 17th century due to European activities. The greater diversity and quantity of ornaments likely resulted from participation in large scale exchange networks that stimulated the circulation of different ornament classes. The frequency of overall ornaments and each subtype declines significantly after European influence fades during the Chinese period in the early 19th century. This may be due to a smaller scale of trading networks, the overall decline of Indigenous populations in Yilan, or the adoption of Han Chinese practices. The decline of the population at Kiwulan may be related to the movement of many Indigenous people southwards to Hualien due to the increasing numbers of Han Chinese immigrants who took over their lands at the end of the 18th century [@Chen2007]. Houses and burials may also be a useful source of evidence to understand population size but a proper treatment of those is beyond the scope of the paper

>>>>

---
title: "Ornaments as indicators of social changes resulting from indirect effects of colonialism in northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      #reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on Indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1350–1850 CE) in northeastern Taiwan show indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered in the residential area. The change reflects an increased use of ornaments and an uneven distribution of ornaments in a pericolonial context. This may hint at increased social inequality resulting from European colonial influence. Trade ornaments give insights into the increased social inequality stimulated by a European colonial presence, and show the agency of Indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; European colonial period; Trade ornaments; Exchange networks; spatial pattern; East Asia; Taiwan
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local Indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

The direct impacts of European colonialism on Indigenous communities in East Asia were much less conspicuous than in island Southeast Asia and Oceania. Direct European colonial rule throughout East Asia was rare and limited, and the question of long-lasting indirect impacts on local Indigenous communities remains largely unanswered. Understanding these indirect effects is important for detecting colonial impacts on Indigenous peoples in the periphery of colonial control [@Acabado2017; @Trabert2017]. In many parts of the world, the introduction of foreign trade goods by colonial traders into local Indigenous societies caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Consumption patterns of foreign goods can give insights into negotiations between colonized and colonizer, and the resistance and accommodations of Indigenous people through their daily cultural practices [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000; @Voss2005]. Northeastern Taiwan is an ideal context to study pericolonial influence because although there was a prominent Spanish and Dutch colonial presence in parts of Taiwan, the northeastern region was isolated from intensive direct contact by the Xueshan Mountains.

This article describes personal ornaments excavated from the upper component of Kiwulan (1350 CE-1850 CE), the largest Iron Age settlement on the Yilan plain in northeastern Taiwan. Ornaments are found at many Iron Age sites in Yilan, but only Kiwulan shows clear stratigraphic contexts from pre-European period to modern time. The first recorded European presence in Yilan was a Spanish revenge attack on Indigenous villages in 1632 [@Borao2001, pp. 163]. In 1647 the Dutch attacked villages and forced them to accept colonial rule and pay an annual tribute without strict enforcement of colonial power that allowed flexibility for Indigenous people in Yilan [@Andrade2007; @Kang2016]. According to Dutch census reports in 1650, Kiwulan was the largest Indigenous settlement in the plain, with a population of 840 adults [@Nakamura1938, pp. 12]. Following defeat of Dutch by the Chinese general Koxinga in 1661-1662, the Dutch abandoned northern Taiwan. Direct contact with Han Chinese is indicated by Qing dynasty census reports mentioning Yilan villages in 1821 [@Yao1996].

@@

Figure \@ref(fig:plot-altogether) shows the distribution of frequencies for subtypes in each major class. Spearman's correlation test shows that there is a significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of the effects of sample size. Carnelian beads and metal rings have greater quantity and variety of shapes compared to copper bells and glass beads during the European period. The greater varieties for carnelian beads and metal rings might indicate multiple origins due to participation in large scale trade networks stimulated by the European presence. In contrast, copper bells have less variety, typically >2 cm long with a wide variety of human faces as a motif. Although glass beads have less variety in size, presenting as small (0.5-1 cm) or medium (1-2 cm), they have a wide variety of colors or patterns mostly made by a winding technique and with high lead content indicating possibly from China [@Cheng2008].

@@

An indirect colonial influence may be indicated at Kiwulan by the greater diversity of ornament types and materials during the European period. Yilan was involved in complex trading networks both on a regional scale with other Indigenous groups and Chinese merchants, and at a global scale with Europeans, including the Dutch and the Spanish. Those trade ornaments have multiple origins, including Southeast Asia and China, and were first introduced into northeastern Taiwan by Chinese merchants before the 17th century. Later, trade activities became more frequent and intense in the 17th century due to European activities. The greater diversity and quantity of ornaments likely resulted from participation in large scale exchange networks that stimulated the circulation of different ornament classes. The frequency of overall ornaments and each subtype declines significantly after European influence fades during the Chinese period in the early 19th century. This may be due to a smaller scale of trading networks, the overall decline of Indigenous populations in Yilan, or the adoption of Han Chinese practices. The decline of the population at Kiwulan may be related to the movement of many Indigenous people southwards to Hualien due to the increasing numbers of Han Chinese immigrants who took over their lands at the end of the 18th century [@Chen2007]. Houses and burials may also be a useful source of evidence to understand population size but a proper treatment of those is beyond the scope of the paper.
